---
title: "TimeSeries Skeleton"
author: "Fanny Chow, Michael McCormack"
date: "March 12, 2018"
output:
  pdf_document: default
  html_document:
    df_print: paged
---
```{r, include = FALSE}
suppressMessages(library(tidyverse))
library(reshape2)
library(ggfortify)
library(forecast)
library(data.table)
library(here)


```
## Step 1: Get the data.  
Download from website https://cdn.rawgit.com/mikejt33/DataViz/246c2026/data/flights.csv.gz
Easiest to unzip locally then read in the data as a csv file (hint: read.table() is typically faster than read.csv)

Easiest to unzip locally then read in the data as a csv file (hint: read.table() is typically faster than read.csv)  
```{r}
flights.path <- here("flights.csv")
flights <- read.table(flights.path, header = TRUE, sep=',')
```
## Step 2: Prepare the data.  

*Are there any null values?

Time series data needs to be over a regular time interval. Calculate the average departure delay time and/or average arrival delay time for each day of 2017.  

Note the number of NA null values.  
```{r}
str(flights)
```


```{r}
# aggregate(DEP_DELAY~DAY_OF_WEEK, data = flights, mean)
# mean.dept.delay <- tapply(flights$FL_DATE, flights$DEP_DELAY, mean, na.rm=TRUE)

mean.dept.delay <- flights %>% 
  arrange(FL_DATE) %>% 
  group_by(factor(FL_DATE)) %>%
  summarise(mean(DEP_DELAY, na.rm = TRUE))

names(mean.dept.delay) <- c('FL_DATE', 'AVE_DELAY')
```

If you like, compare average delay times for different carriers or different airports by creating multiple time series.


## Step 3: Create a ts object of the data.  
Refer to the slides for tips on how to do this.
```{r}
mean.dept.delay.ts <- ts(mean.dept.delay$AVE_DELAY)
# str(mean.dept.delay)
```

## Step 4: Plot the time series using base package and ggplot (advanced).
Create a basic visualization of the time series  

```{r}
plot(mean.dept.delay.ts, main="Avg Departure Delay for 2017", ylab="avg departure delay (min.)")
```

## Step 5: Smooth the data to reduce noise and identify trends.

### Create your own simple moving average for monthly data. Plot the smoothed data using base package. Plot both the originial and the smoothed data ggplot (advanced). 

__Hints__  
* good StackOverflow reference for moving average in R: https://stackoverflow.com/questions/743812/calculating-moving-average
* watch out for functions that may have been masked by other packages
* ggplot: may need to convert data to long format to plot mutliple series  

Let's say there are 31 days per month for the window size.  
1/n creates intervals. So say if you had quarterly data, we would bin it into 1/4 reps.  
```{r}
# n = window size
ma <- function(x,n=5){stats::filter(x,rep(1/n,n), sides=2)}

moving.average <- ma(mean.dept.delay.ts, 31)

plot(moving.average, ylab="Moving Average", main="Moving Average of Departure Times in 2017")

autoplot(moving.average)


```


```{r}
# generate dates to make plots more legible
dates <- seq(as.Date('2017-01-01'), as.Date('2017-12-31'), by = 'day') # create sequence of dates
plot.ts <- data.frame(smoothed_MA = moving.average, ave_delay=as.numeric(mean.dept.delay.ts),date=dates)

```

```{r}
# melt into long format 
# plot both the original and smoothed data can be easily plotted on the same axes
plot.ts.melted <- melt(plot.ts, id='date')

ggplot(plot.ts.melted) + geom_line(aes(x = date, y  = value, col = variable)) + labs(x = 'Date', y = 'Average arrival delay (minutes)', title='Smoothed vs MA Delay') + theme_minimal() 

```

__Questions__  
1. How does the neighborhood size, i.e. the number of points in each localized subset, affect the amount of smoothing?   

When you increase the window size, you have more smoothing.  

2. What happened to endpoints of the smoothed data?  
You lose the first and last endpoints. You cannot intialize a subset of calculating the MA at those points, so they're set to NA.  

### Advanced: Smooth the same data using Local Regression (loess). Plot smoothed data using base package. Plot all three series (original, smoothed by MA, and smoothed by loess) using ggplot (advanced).

__Hint__  
* loess() librarys all predictors to be numerical so dates cannot be used  

Try different values for the span argument and see how it affects the amount of smoothing.  

```{r}
# names(mean.dept.delay) <- c('FL_DATE', 'average.delay')

# lowess.index <- seq(1:nrow(mean.dept.delay))
mean.dept.delay$index <- 1:nrow(mean.dept.delay)

# mean.dept.delay <- cbind(mean.dept.delay, lowess.index)
                    
loess.model <- loess(AVE_DELAY~index, data=mean.dept.delay, span = 1/31)

```

```{r}
# smooth the data by using the predict() function
trend.loess <- predict(loess.model)

# plot the smoothed data using base package
plot(trend.loess)
```

```{r}
# plot using ggplot

# creating data frame with dates to improve readability of plots
plot_ts <- cbind(plot.ts, smoothed.loess = trend.loess)

# melt into long format so both the origininal and smoothed data can be plotted on the same axes
plot.ts.melted <- melt(plot.ts, id='date')

ggplot(plot.ts.melted) + geom_line(aes(x = date, y  = value, col = variable)) + labs(x = 'Date', y = 'Average arrival delay (minutes)') 
```





# Dive in Deeper to TimeSeries  
For this portion of our lab we will be using data from the AirPassengers Dataset  

```{r}
data(AirPassengers)
```

## Step 6: Make an inital TimeSeries Visual of the data  
```{r}
timeseries_air = AirPassengers
plot(as.ts(timeseries_air))
```

## Step 7: Compute the Moving Average of this data using forecast package and vizualize this  
```{r}
trend_air = ma(timeseries_air)
               # , order = 12, centre = T)
#lines(trend_air)
plot(as.ts(trend_air))
```
## Step 8: Remove the Trend from the data and Visualize this  
```{r}
detrend_air = timeseries_air / trend_air
plot(as.ts(detrend_air))
```
## Step 9: Create a decomposition of the data by month  
-- Hint (Frequency = 12)
```{r}
ts_air = ts(timeseries_air, frequency = 12)
decompose_air = decompose(ts_air, "multiplicative")
 
plot(decompose_air) 

### Advanced answers
plot(as.ts(decompose_air$seasonal))
plot(as.ts(decompose_air$trend))
plot(as.ts(decompose_air$random))
```





