---
title: "TimeSeries Skeleton"
author: "Fanny Chow"
date: "March 12, 2018"
output:
  html_document:
    df_print: paged
---
```{r, include = FALSE}
suppressMessages(library(tidyverse))
library(reshape2)
library(ggfortify)
library(forecast)
```
## Step 1: Get the data.  
Download from website https://cdn.rawgit.com/mikejt33/DataViz/246c2026/data/flights.csv.gz
Easiest to unzip locally then read in the data as a csv file (hint: read.table() is typically faster than read.csv)  
```{r}
library(data.table)
library(here)
flights.path <- here("flights.csv")
flights <- read.table(flights.path, header = TRUE, sep=',')
```
## Step 2: Prepare the data.  

*Are there any null values?

Time series data needs to be over a regular time interval. Calculate the average departure delay time and/or average arrival delay time for each day of 2017.


```{r}
# aggregate(DEP_DELAY~DAY_OF_WEEK, data = flights, mean)
# mean.dept.delay <- tapply(flights$FL_DATE, flights$DEP_DELAY, mean, na.rm=TRUE)

mean.dept.delay <- flights %>% group_by(factor(FL_DATE)) %>%
   summarize(average.delay = mean(DEP_DELAY, na.rm = TRUE))

```

If you like, compare average delay times for different carriers or different airports by creating multiple time series.


## Step 3: Create a ts object of the data.  
Refer to the slides for tips on how to do this.
```{r}
mean.dept.delay <- ts(mean.dept.delay$average.delay)
str(mean.dept.delay)
```

## Step 4: Plot the time series using base package and ggplot (advanced).
Create a basic visualization of the time series  

```{r}
plot(mean.dept.delay, main="Avg Departure Delay for 2017", ylab="avg departure delay (min.)")
```

## Step 5: Smooth the data to reduce noise and identify trends.

### Create your own simple moving average for monthly data. Plot the smoothed data using base package. Plot both the originial and the smoothed data ggplot (advanced). 

__Hints__  
* good StackOverflow reference for moving average in R: https://stackoverflow.com/questions/743812/calculating-moving-average
* watch out for functions that may have been masked by other packages
* ggplot: may need to convert data to long format to plot mutliple series  

Let's say there are 31 days per month for the window size.  
1/n creates intervals. So say if you had quarterly data, we would bin it into 1/4 reps.  
```{r}
# n = window size
ma <- function(x,n=5){stats::filter(x,rep(1/n,n), sides=2)}

moving.average <- ma(mean.dept.delay$average.delay, 31)
plot(moving.average, ylab="Moving Average", main="Moving Average of Departure Times in 2017")
# autoplot(moving.average)

# apply(mean.dept.delay, ma)


```

__Questions__  
1. How does the neighborhood size, i.e. the number of points in each localized subset, affect the amount of smoothing?  
2. What happened to endpoints of the smoothed data?  


### Advanced: Smooth the same data using Local Regression (loess). Plot smoothed data using base package. Plot all three series (original, smoothed by MA, and smoothed by loess) using ggplot (advanced).

__Hint__  
* loess() librarys all predictors to be numerical so dates cannot be used  

Try different values for the span argument and see how it affects the amount of smoothing.  

```{r}
names(mean.dept.delay) <- c('FL_DATE', 'average.delay')
lowess.index <- seq(1:nrow(mean.dept.delay))
mean.dept.delay <- cbind(mean.dept.delay, lowess.index)
                    
loess(average.delay~lowess.index, data=mean.dept.delay, span = 1/31)
```


# Dive in Deeper to TimeSeries  

For this portion of our lab we will be using data from the AirPassengers Dataset  

```{r}
data(AirPassengers)
```

## Step 6: Make an inital TimeSeries Visual of the data  
```{r}

```
## Step 7: Compute the Moving Average of this data using forecast package and vizualize this  
```{r}

```
## Step 8: Remove the Trend from the data and Visualize this  
```{r}

```
## Step 9: Create a decomposition of the data by month  
-- Hint (Frequency = 12)
```{r}

```




