---
title: "TimeSeries Skeleton"
author: "Michael McCormack"
date: "March 12, 2018"
output: pdf_document

---

require(tidyverse)
require(ggfortify)
require(forecast)


## Step 1: Get the data
Download from website https://cdn.rawgit.com/mikejt33/DataViz/246c2026/data/flights.csv.gz
Easiest to unzip locally then read in the data as a csv file.
```{r}
flights <- read.csv('flights.csv')
```
## Step 2: Initial EDA
Are there any null values?

Isolate the data you want to use.  We recommend using arrival delay and departure delay times.  Or Origin/Destination cities.  Or Aggregate by Carrier


## Step 3: Fit the data to a TimeSeries object.
Refer to the slides for tips on how to do this.


## Step 4: Use TimeSeries object data to make an initial Visualization  
Create a basic Visualization of the TimeSeries object data

###Advanced 
See if you can figure out how to do this in ggplot
-hint the function is called autoplot and is part of ggfortify.



# Dive in Deeper to TimeSeries
For this portion of our lab we will be using data from the AirPassengers Dataset

```{r}
data(AirPassengers)
```
## Step 5: Make an inital TimeSeries Visual of the AirPassenger Data

## Step 6: Compute the Moving Average of this data and visualize this

## Step 7: Remove the Trend from the data and visualize this

## Step 8: Create a decomposition of the data by month
-- Hint (Frequency = 12), use the decompose function,  and remember this is multiplicative

###Advanced: Visually isolate certain parts of the graph
How would you only represent the random graph.

## Step 9: Create your own simple moving average for monthly data
Hint - A good starting place is StackOverFlow
https://stackoverflow.com/questions/743812/calculating-moving-average



