---
title: "TimeSeries"
author: "Michael McCormack"
date: "March 12, 2018"
output:
  html_document:
    df_print: paged
---
```{r}
suppressMessages(require(tidyverse))
require(reshape2)
require(ggfortify)
require(forecast)
```
## Step 1: Get the data.  
Download from website https://cdn.rawgit.com/mikejt33/DataViz/246c2026/data/flights.csv.gz
Easiest to unzip locally then read in the data as a csv file (hint: read.table() is typically faster than read.csv)  
```{r}
flights <- read.table('flights.csv', header = TRUE, sep = ',')
```
## Step 2: Prepare the data.  

*Are there any null values?

Time series data needs to be over a regular time interval. Calculate the average departure delay time and/or average arrival delay time for each day of 2017.


```{r}
arr_delay <- flights %>% arrange(FL_DATE) %>% group_by(FL_DATE) %>% summarise(mean(ARR_DELAY, na.rm = TRUE))
names(arr_delay) <- c('FL_DATE', 'AVE_DELAY')

dep_delay <- flights %>% arrange(FL_DATE) %>% group_by(FL_DATE) %>% summarise(mean(DEP_DELAY, na.rm = TRUE))
names(dep_delay) <- c('FL_DATE', 'AVE_DELAY')
```

If you like, compare average delay times for different carriers or different airports by creating multiple time series.


## Step 3: Create a ts object of the data.  
Refer to the slides for tips on how to do this.
```{r}
arr_delay_ts <- ts(arr_delay$AVE_DELAY)
dep_delay_ts <- ts(dep_delay$AVE_DELAY)
```

## Step 4: Plot the time series using base package and ggplot (advanced).
Create a basic visualization of the time ser

```{r}
plot(arr_delay_ts)
plot(dep_delay_ts)

# Advanced Portion, requires ggfortify
autoplot(arr_delay_ts)
```

## Step 5: Smooth the data to reduce noise and identify trends.

### Create your own simple moving average for monthly data. Plot the smoothed data using base package. Plot both the originial and the smoothed data ggplot (advanced). 

__Hints__  
* good StackOverflow reference for moving average in R: https://stackoverflow.com/questions/743812/calculating-moving-average
* watch out for functions that may have been masked by other packages
* ggplot: may need to convert data to long format to plot mutliple series  

```{r}
# simple moving average for monthly data, n = is neighborhood size i.e. the number of point in each localized subset
moving_ave <- function(x,n){stats::filter(x,rep(1/n,n), sides=2)}
trend_ma <- moving_ave(arr_delay_ts,31)

# plot smoothed data using base package
plot(trend_ma)


# plot using ggplot
# autoplot, requires ggfortify
autoplot(trend_ma, ts.color = 'blue')


# creating data frame with dates to improve readability of plots
dates <- seq(as.Date('2017-01-01'), as.Date('2017-12-31'), by = 'day') # create sequence of dates
plot_ts <- data.frame(smoothed_MA = trend_ma, ave_delay=as.numeric(arr_delay_ts),date=dates)

# melt into long format so both the origininal and smoothed data can be easily plotted on the same axes
plot_ts_melted <- melt(plot_ts, id='date')

ggplot(plot_ts_melted) + geom_line(aes(x = date, y  = value, col = variable)) + labs(x = 'Date', y = 'Average arrival delay (minutes)') 
```

__Questions__  
1. How does the neighborhood size, i.e. the number of points in each localized subset, affect the amount of smoothing?  
2. What happened to endpoints of the smoothed data?  


__Answers__
1. Increasing window size increases the amount of smoothing.
2. The first and last (n-1)/2 data points are lost, where n is odd and denotes the neighborhood size. The moving average at those points in time don't have access to a full localized subset of size n and are set as NA's.


### Advanced: Smooth the same data using Local Regression (loess). Plot smoothed data using base package. Plot all three series (original, smoothed by MA, and smoothed by loess) using ggplot (advanced).

__Hint__  
* loess() requires all predictors to be numerical so dates cannot be used  

Try different values for the span argument and see how it affects the amount of smoothing.  
```{r}
# create index variable
arr_delay$index <- 1:nrow(arr_delay)  
# get the loess model, span is the proportion of data used in each localized subset
loess_model <- loess(AVE_DELAY ~ index, data=arr_delay, span=0.25)
# smooth the data by using the predict() function
trend_loess <- predict(loess_model)

# plot the smoothed data using base package
plot(trend_loess)


# plot using ggplot

# creating data frame with dates to improve readability of plots
plot_ts <- cbind(plot_ts, smoothed_loess = trend_loess)

# melt into long format so both the origininal and smoothed data can be plotted on the same axes
plot_ts_melted <- melt(plot_ts, id='date')

ggplot(plot_ts_melted) + geom_line(aes(x = date, y  = value, col = variable)) + labs(x = 'Date', y = 'Average arrival delay (minutes)') 
```


# Dive in Deeper to TimeSeries  
For this portion of our lab we will be using data from the AirPassengers Dataset
```{r}
data(AirPassengers)
```
## Step 6: Make an inital TimeSeries Visual of the data  
```{r}
timeseries_air = AirPassengers
plot(as.ts(timeseries_air))
```
## Step 7: Compute the Moving Average of this data using forecast package and vizualize this  
```{r}
trend_air = ma(timeseries_air, order = 12, centre = T)
#lines(trend_air)
plot(as.ts(trend_air))
```
## Step 8: Remove the Trend from the data and Visualize this  
```{r}
detrend_air = timeseries_air / trend_air
plot(as.ts(detrend_air))
```
## Step 9: Create a decomposition of the data by month  
-- Hint (Frequency = 12)
```{r}
ts_air = ts(timeseries_air, frequency = 12)
decompose_air = decompose(ts_air, "multiplicative")
 
plot(decompose_air) 

### Advanced answers
plot(as.ts(decompose_air$seasonal))
plot(as.ts(decompose_air$trend))
plot(as.ts(decompose_air$random))
```

